---
title: "In-class Exercise 2: Spatial Point Patterns Analysis"
author: "Christover Manafe"
date: "2024-09-02"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
format: 
  html:
    code-fold: true
    code-summary: "code chunk"
    number-sections: true
    number-depth: 4
---

# Installing and loading R packages

## Install **maptools** package

**maptools** is retired and its binary is removed from CRAN.  However, we can download it from [Posit Public Package Manager](https://packagemanager.posit.co/client/#/repos/cran/setup?snapshot=2023-10-13&distribution=windows&r_environment=ide) snapshots by using following code

```{r}
#| eval: false
install.packages("maptools",
                 repos = "https://packagemanager.posit.co/cran/2023-10-13")
```

> Include `#| eval: false` in the code chunk to avoid **maptools** being downloaded and installed repetitively every time the Quarto document is rendered.

## Install remaining packages

The code chunk below install and load following packages into R environment:

-   [**sf**](https://r-spatial.github.io/sf/), a relatively new R package specially designed to import, manage and process vector-based geospatial data in R.
-   [**spatstat**](https://spatstat.org/), which has a wide range of useful functions for point pattern analysis. In this hands-on exercise, it will be used to perform 1st- and 2nd-order spatial point patterns analysis and derive kernel density estimation (KDE) layer.
-   [**raster**](https://cran.r-project.org/web/packages/raster/) which reads, writes, manipulates, analyses and model of gridded spatial data (i.e. raster). In this hands-on exercise, it will be used to convert image output generate by spatstat into raster format.
-   [**maptools**](https://cran.r-project.org/web/packages/maptools/index.html) which provides a set of tools for manipulating geographic data. In this hands-on exercise, we mainly use it to convert *Spatial* objects into *ppp* format of **spatstat**.
-   [**tmap**](https://cran.r-project.org/web/packages/tmap/index.html) which provides functions for plotting cartographic quality static point patterns maps or interactive maps by using [leaflet](https://leafletjs.com/) API.

```{r}
pacman::p_load(sf, raster, spatstat, tmap, tidyverse)
```

# The data

Data sets in this exercise are as follows:

-   `CHILDCARE` (Point Feature Data)
    -   This dataset provides both location and attribute information of childcare centres across Singapore.
    -   Format: GeoJSON
    -   Source: [data.gov.sg](https://data.gov.sg/)
-   `MP14_SUBZONE_WEB_PL` (Polygon Feature Data)
    -   This dataset provide information of URA 2014 Master Plan Planning Subzone boundary data.
    -   Format: ESRI Shapefile.
    -   Source: [data.gov.sg](https://data.gov.sg/)
-   `CostalOutline` (Polygon Feature Data)
    -   This dataset shows the national boundary of Singapore.
    -   Format: ESRI Shapefile
    -   Source: [Singapore Land Authority](https://www.sla.gov.sg/)


We will use [st_read()](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package will be used to import these three geospatial data sets into R.

::: panel-tabset
## Childcare data

Since the `childcare_sf` simple feature data frame is in the WGS84 geodetic CRS, which is not ideal for geospatial analysis, the `st_transform()` function from the **sf** package is used to reproject the data to the SVY21 coordinate system during import.

```{r}
childcare_sf <- st_read("data/child-care-services-geojson.geojson") %>%
  st_transform(crs = 3414)
```

Let's verify the crs of the data frame to ensure we're using EPSG 3414.

```{r}
st_crs(childcare_sf)
```

## Coastal outline data

Import coastal outline data using `st_read()` function

```{r}
sg_sf <- st_read(dsn = "data", layer="CostalOutline")
```

Let's check coordinate system of this data frame

```{r}
st_crs(sg_sf)
```

Coastal outline data frame is using EPSG 9001 instead of 3414 which is suitable for CRS SVY21. Let's assign correct EPSG code using `st_set_crs()` then verify the output.

```{r}
sg_sf = st_set_crs(sg_sf, 3414)

st_crs(sg_sf)
```

## Master Plan 2014 data

Let's load the Master Plan Planning data using `st_read()` function

```{r}
mpsz_sf <- st_read(dsn = "data", 
                layer = "MP14_SUBZONE_WEB_PL")
```

Let's check coordinate system of this data frame

```{r}
st_crs(mpsz_sf)
```

`mpsz_sf` is also using EPSG 9001 instead of 3414 which is suitable for CRS SVY21. Let's assign correct EPSG code using `st_set_crs()` then verify the output.

```{r}
mpsz_sf <- st_set_crs(mpsz_sf,3414)
st_crs(mpsz_sf)
```
:::


## working with `st_union()`
is used to derive the coastal outline sf tibble data.frame

```{r}
sg_sf <- mpsz_sf %>%
  st_union()
```

*sg_sf* will look similar to the figure below.

```{r}
ggplot() +
  geom_sf(data = sg_sf)
```


# Monte Carlo Simulation

In order to ensure reproducibility, it is important to include the code chunk below before using **spatstat** functions involve Monte Carlo simulation

```{r}
set.seed(2024)
```

# Importing Traffic Accident Data

```{r}
rdacc_sf <- read_csv("data/thai_road_accident_2019_2022.csv") %>%
  filter(!is.na(longitude) & longitude != "",
         !is.na(latitude) & latitude != "") %>%
  st_as_sf(coords = c(
    "longitude", "latitude"),
    crs=4326) %>%
  st_transform(crs = 32647)
```

# Visualising the accident data

```{r}
rdacc_ppp <- as.ppp(rdacc_sf)

plot(rdacc_ppp)
```