---
title: "In-class Exercise 9: Modelling Geographical Accessibility"
author: "Christover Manafe"
date: "2024-10-28"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
format: 
  html:
    code-fold: false
    code-summary: "code chunk"
    number-sections: true
    number-depth: 4
---

# The Packages

We'll load following packages in R environment.

```{r}
pacman::p_load(tmap, SpatialAcc, sf, ggstatsplot, reshape2, tidyverse)
```

# The Data

Importing geospatial data and aspatial data. We'll use the same data as those in `Hands-on Exercise 9`.

::: panel-tabset
## Geospatial

```{r}

mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_NO_SEA_PL") %>%
  st_transform(3414)
hexagons <- st_read(dsn = "data/geospatial", layer = "hexagons") %>%
  st_transform(3414)
eldercare <- st_read(dsn = "data/geospatial", layer = "ELDERCARE") %>%
  st_transform(3414)
```

## Aspatial

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", skip = 0)
```
:::

# Data cleaning and updating attributes

## Supply

Let's use only `fid` and `ADDRESSPOS` from `eldercare` data frame, and we'll add new column `capacity` with value `100` for simplicity. In real life analysis, we would need to survey these eldercare facilities to find the actual capacity.

```{r}
eldercare <- eldercare%>%
  select(fid, ADDRESSPOS) %>%
  mutate(capacity = 100)
```

## Demand

Let's use only `fid` from `hexagons` data frame, and we'll add new column `demand` with value `100` for simplicity.

```{r}
hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```

## OD Matrix

```{r}
distmat <- ODMatrix %>%
  select(origin_id, destination_id, total_cost) %>%
  spread(destination_id, total_cost) %>%
  select(c(-c('origin_id')))
```

Let's create a large matrix from `distmat` and also convert the unit of measurement from meters to kilometers.

```{r}
distmat_km <- as.matrix(distmat/1000)
```

# Computing Hansen's Accessibility

```{r}
acc_hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km,
                            d0 = 50,
                            power = 2,
                            family = "Hansen"))
colnames(acc_hansen) <- "accHansen"
acc_hansen <- as_tibble(acc_hansen)
hexagon_hansen <- bind_cols(hexagons, acc_hansen)
```

# Visualising Accesibility

```{r fig.height=8, fig.width=12}
mapex <- st_bbox(hexagons)
tmap_mode("plot")
tm_shape(hexagon_hansen, bbox = mapex) +
  tm_fill(col = "accHansen",
          n = 10,
          style = "quantile",
          border.col = "block",
          border.lwd = 1) +
  tm_shape(eldercare) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to eldercare: Hansen method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

# Statistical Graphic

```{r fig.height=8, fig.width=12}
hexagon_hansen <- st_join(hexagon_hansen, mpsz, join = st_intersects)
ggbetweenstats(
  data = hexagon_hansen,
  x = REGION_N,
  y = accHansen,
  type = "p"
)
```
